---
title: "r4ds_5"
author: "Vuk Dinic"
date: "2025-02-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Data transformation

## 5.1 Introduction
### 5.1.1 Prerequisites
```{r}
library(nycflights13)
library(tidyverse)
```
### 5.1.2 nycflights13
```{r}
flights <- nycflights13::flights
```

## 5.2 Filter rows with filter()
```{r}
filter(flights, month == 1, day == 1)
```

```{r}
jan1 <- filter(flights, month == 1, day == 1)
```

```{r}
(dec25 <- filter(flights, month == 12, day == 25))
```

### 5.2.1 Comparisons
```{r}
filter(flights, month = 1)
```

```{r}
sqrt(2) ^ 2 == 2
```

```{r}
1 / 49 * 49 == 1
```

```{r}
near(sqrt(2) ^ 2,  2)
```

```{r}
near(1 / 49 * 49, 1)
```
### 5.2.2 Logical operators
```{r}
filter(flights, month == 11 | month == 12)
```

```{r}
nov_dec <- filter(flights, month %in% c(11, 12))
```

```{r}
filter(flights, !(arr_delay > 120 | dep_delay > 120))
filter(flights, arr_delay <= 120, dep_delay <= 120)
```

### 5.2.3 Missing values
```{r}
NA > 5
```

## 5.3 Arrange rows with arrange()
```{r}
arrange(flights, year, month, day)
```

```{r}
arrange(flights, desc(dep_delay))
```

```{r}
df <- tibble(x = c(5, 2, NA))
```

## 5.4 Select columns with select()
```{r}
# Select columns by name
select(flights, year, month, day)
```

```{r}
rename(flights, tail_num = tailnum)
```

```{r}
select(flights, time_hour, air_time, everything())
```

## 5.5 Add new variables with mutate()
```{r}
flights_sml <- select(flights, 
  year:day, 
  ends_with("delay"), 
  distance, 
  air_time
)

flights_sml <- mutate(flights_sml,
  gain = dep_delay - arr_delay,
  speed = distance / air_time * 60
)
```

```{r}
mutate(flights_sml,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

```{r}
transmute(flights,
  gain = dep_delay - arr_delay,
  hours = air_time / 60,
  gain_per_hour = gain / hours
)
```

### 5.5.1 Useful creation functions
```{r}
transmute(flights,
  dep_time,
  hour = dep_time %/% 100,
  minute = dep_time %% 100
)
```

```{r}
(x <- 1:10)
```
```{r}
lag(x)
```
```{r}
lead(x)
```
## 5.6 Grouped summaries with summarise()
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

```{r}
by_day <- group_by(flights, year, month, day)
```
```{r}
summarise(by_day, delay = mean(dep_delay, na.rm = TRUE))
```

### 5.6.1 Combining multiple operations with the pipe
```{r}
by_dest <- group_by(flights, dest)
```
```{r}
delay <- summarise(by_dest,
  count = n(),
  dist = mean(distance, na.rm = TRUE),
  delay = mean(arr_delay, na.rm = TRUE)
)
```
```{r}
delay <- filter(delay, count > 20, dest != "HNL")
```

```{r}
ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)
```

```{r}
delays <- flights %>% 
  group_by(dest) %>% 
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>% 
  filter(count > 20, dest != "HNL")
```

### 5.6.2 Missing values
```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

```{r}
flights %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay, na.rm = TRUE))
```

```{r}
not_cancelled <- flights %>% 
  filter(!is.na(dep_delay), !is.na(arr_delay))
```

```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(mean = mean(dep_delay))
```

### 5.6.3 Counts
```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay)
  )
```
```{r}
ggplot(data = delays, mapping = aes(x = delay)) + 
  geom_freqpoly(binwidth = 10)
```

```{r}
delays <- not_cancelled %>% 
  group_by(tailnum) %>% 
  summarise(
    delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
```
```{r}
ggplot(data = delays, mapping = aes(x = n, y = delay)) + 
  geom_point(alpha = 1/10)
```

```{r}
delays %>% 
  filter(n > 25) %>% 
  ggplot(mapping = aes(x = n, y = delay)) + 
    geom_point(alpha = 1/10
```

### 5.6.4 Useful summary functions
```{r}
not_cancelled %>% 
  group_by(year, month, day) %>% 
  summarise(
    avg_delay1 = mean(arr_delay),
    avg_delay2 = mean(arr_delay[arr_delay > 0]) # the average positive delay
  )
```

### 5.6.5 Grouping by multiple variables
```{r}

```

### 5.6.6 Ungrouping
```{r}

```

## 5.7 Grouped mutates (and filters)
```{r}
flights_sml %>% 
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)
```

