---
title: "r4ds_1"
author: "Vuk Dinic"
date: "2025-02-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1.1 Prerequisites
```{r tidiverse}
library(tidyverse)
```

```{r penguines, gg themes}
library(palmerpenguins)
library(ggthemes)
```

# 1.2 First steps
```{r The penguins data frame}
penguins
```

```{r See all variables}
glimpse(penguins)
```
# 1.2.3 Creating a ggplot
```{r Blank plot}
ggplot(data = penguins)
```
```{r Aesthetics}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
)
```

```{r First plot}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point()
```

# 1.2.4 Adding aesthetics and layers
```{r Colors on plot}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point()
```
```{r Adding lines}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = species)
) +
  geom_point() +
  geom_smooth(method = "lm")
```
```{r One line}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species)) +
  geom_smooth(method = "lm")
```
```{r Adding shapes}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(mapping = aes(color = species, shape = species)) +
  geom_smooth(method = "lm")
```
```{r Final plot}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(aes(color = species, shape = species)) +
  geom_smooth(method = "lm") +
  labs(
    title = "Body mass and flipper length",
    subtitle = "Dimensions for Adelie, Chinstrap, and Gentoo Penguins",
    x = "Flipper length (mm)", y = "Body mass (g)",
    color = "Species", shape = "Species"
  )+
  scale_color_colorblind()
```

# 1.2.5 Exercises
```{r How many rows are in penguins? How many columns?}
dim(penguins)
```

```{r What does the bill_depth_mm variable in the penguins data frame describe? }
?penguins
```

```{r Make a scatterplot of bill_depth_mm vs. bill_length_mm}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm)
)+
  geom_point()
```
```{r What happens if you make a scatterplot of species vs. bill_depth_mm? }
ggplot(
  data = penguins,
  mapping = aes(x = species, y = bill_depth_mm)
)+
  geom_boxplot()
```
na.rm remove missing values
```{r na.rm}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm, color = species)
)+
  geom_point(na.rm = TRUE)
```

```{r Add the caption}
ggplot(
  data = penguins,
  mapping = aes(x = bill_length_mm, y = bill_depth_mm, color = species)
)+
  geom_point(na.rm = TRUE)+
  labs(caption = "Data come from the palmerpenguins package.")
```

```{r Recreate the visualization}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point(na.rm = TRUE, aes(color = bill_depth_mm)) +
  geom_smooth(method = "loess")
```
```{r Run this code first in your head}
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g, color = island)
) +
  geom_point() +
  geom_smooth(se = FALSE)
```
```{r Will these two graphs look different? }
ggplot(
  data = penguins,
  mapping = aes(x = flipper_length_mm, y = body_mass_g)
) +
  geom_point() +
  geom_smooth()

ggplot() +
  geom_point(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  ) +
  geom_smooth(
    data = penguins,
    mapping = aes(x = flipper_length_mm, y = body_mass_g)
  )
```
# 1.3ggplot2 calls
```{r More concise version of code}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```

```{r Pipe version}
penguins |> 
  ggplot(aes(x = flipper_length_mm, y = body_mass_g)) + 
  geom_point()
```

# 1.4 Visualizing distributions
# 1.4.1 A categorical variable
```{r A categorical variable}
ggplot(penguins, aes(x = species)) +
  geom_bar()
```

```{r Ordered Bear plot}
ggplot(penguins, aes(x = fct_infreq(species))) +
geom_bar()
```

## 1.4.2 A numerical variable
```{r Histogram}
ggplot(penguins, aes(x = body_mass_g, fill = species)) +
  geom_histogram(binwidth =  200, color = "black")
```

```{r Line plot}
ggplot(penguins, aes(x = body_mass_g, color = species)) +
  geom_density()
```
# 1.4.3 Exercises
```{r Make a bar plot of species of penguins} 
ggplot(penguins, aes(y = species)) +
  geom_bar()
```

```{r How are the following two plots different? }
ggplot(penguins, aes(x = species)) +
  geom_bar(color = "red")

ggplot(penguins, aes(x = species)) +
  geom_bar(fill = "red")
```

```{r Make a histogram of the carat variable in the diamonds}
ggplot(diamonds, aes(x=carat)) +
  geom_histogram(bins = 9)
```

# 1.5Visualizing relationships

## 1.5.1 A numerical and a categorical variable
```{r Boxplot}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot()
```

```{r Line plot 2}
ggplot(penguins, aes(x=body_mass_g, color = species)) +
  geom_density(linewidth = 0.75)
```

```{r Line plot with the fill}
ggplot(penguins, aes(x = body_mass_g, color = species, fill = species)) +
  geom_density(alpha = 0.5)
```

# 1.5.2 Two categorical variables
```{r Two categorical variables}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar()
```

```{r Relative frequency plot}
ggplot(penguins, aes(x = island, fill = species)) + 
  geom_bar(position = "fill")
```

# 1.5.3 Two numerical variables
```{r Scatterplot}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
```
# 1.5.4 Three or more variables
```{r Three or more variables scatter}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = island))
```
```{r facets}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, shape = species))+
  facet_wrap(~island)
```
# 1.5.5 Exercies
```{r mpg The mpg data frame}
mpg
```

```{r Structure of mpg}
str(mpg)
```
Which variables categorical? manufacturer, model, trans, drv, fl, class
Which variables are numerical? displ, year, cyl, cty, hwy

```{r Scatter mpg}
ggplot(mpg, aes(x = displ, y = hwy, color = cty)) +
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy, size = cty)) +
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy, color = cty, size = cty)) +
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy, shape = as.factor(cty))) +
  geom_point()

ggplot(mpg, aes(x = displ, y = hwy, linewidth = cty)) +
  geom_point()
```

```{r Make a scatterplot of bill_depth_mm vs. bill_length_mm and color the points by species}
ggplot(penguins, aes(x=bill_depth_mm, y=bill_length_mm, color=species))+
  geom_point()
```

```{r Why does the following yield two separate legends?}
ggplot(
  data = penguins,
  mapping = aes(
    x = bill_length_mm, y = bill_depth_mm, 
    color = species, shape = species
  )
) +
  geom_point() +
  labs(color = "Species")
```

```{r Create the two following stacked bar plots}
ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(position = "fill")
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(position = "fill")
```
# 1.6 Saving your plots
```{r Save plot}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point()
ggsave(filename = "penguin-plot.png")
```
# 1.6.1 Exercises
```{r Save two plots with the same name}
ggplot(mpg, aes(x = class)) +
  geom_bar()
ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave("mpg-plot.png")
```

