---
title: "r4ds_02"
author: "Vuk Dinic"
date: "2025-02-17"
output: html_document
---
# Workflow: basics
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 2.1 Coding basics
```{r Bacis calculations}
1 / 200 * 30

(59 + 73 + 2) / 3

sin(pi / 2)
```

```{r Create new object}
x <- 3 * 4
x
```
```{r combining elements into a vector}
primes <- c(2, 3, 5, 7, 11, 13)
```

```{r Basic arithmetic on vectors}
primes * 2
```
Alt plus minus sign creates <-

# 2.2 Comments
```{r Commenting}
# create vector of primes
primes <- c(2, 3, 5, 7, 11, 13)

# multiply primes by 2
primes * 2
#> [1]  4  6 10 14 22 26
```

# 2.3 What’s in a name?
```{r Object names}
this_is_a_really_long_name <- 2.5
```

# 2.4 Calling functions
```{r you can use TAB to autocomplete your code}
seq(from = 1, to = 10)
```

# 2.5 Exercises
```{r Why this code doesn't work?}
my_variable <- 10
my_varıable
```
my_variable and my_varıable are not the same. Take a look at i

```{r Tweak each of the following R commands so that they run correctly}
libary(tidyverse) #should be library and tidyverse

ggplot(dTA = mpg) + 
  geom_point(maping = aes(x = displ y = hwy)) + #you should add , after disply
  geom_smooth(method = "lm ) " #you should add " after lm
```

```{r Which of the two plots is saved as mpg-plot.png?}
my_bar_plot <- ggplot(mpg, aes(x = class)) +
  geom_bar()
my_scatter_plot <- ggplot(mpg, aes(x = cty, y = hwy)) +
  geom_point()
ggsave(filename = "mpg-plot.png", plot = my_bar_plot)
```
