---
r4ds_12title: "r4ds_12"
author: "Vuk Dinic"
date: "2025-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 12 Logical vestors
## 12.1 Introduction
### 12.1.1 Prerequsites
```{r}
library(tidyverse)
library(nycflights13)
```

```{r}
x <- c(1, 2, 3, 5, 7, 11, 13)
```
```{r}
x * 2
```
```{r}
df <- tibble(x)
```
```{r}
df |> mutate(y = x * 2)
```
## 12.2 Comparisons

```{r}
flights |> 
  filter(dep_time > 600 & dep_time < 2000 & abs(arr_delay) < 20)
```
```{r}
flights |> 
  mutate(
    daytime = dep_time > 600 & dep_time < 2000,
    approx_ontime = abs(arr_delay) < 20,
    .keep = "used"
  )
```
```{r}
flights |> 
  mutate(
    daytime = dep_time > 600 & dep_time < 2000,
    approx_ontime = abs(arr_delay) < 20,
  ) |> 
  filter(daytime & approx_ontime)
```
### 1.2.1 Floating point comarison

```{r}
x <- c(1 / 49 * 49, sqrt(2) ^ 2)
x
```
```{r}
x == c(1,2)
```
```{r}
print(x, digits = 16)
```
```{r}
near(x, c(1, 2))
```

### 12.2.2 Missing values
```{r}
NA > 5
```
```{r}
10 == NA
```
```{r}
NA = NA
```
```{r}
# We don't know how old Mary is
age_mary <- NA
```
```{r}
# We don't know how old John is
age_john <- NA
```

```{r}
# Are Mary and John the same age?
age_mary == age_john
```
We don't know!

```{r}
flights |> 
  filter(dep_time == NA)
```
### 12.2.3 is.na()
```{r}
is.na(c(TRUE, NA, FALSE))
```
```{r}
is.na(c(1, NA, 3))
```
```{r}
is.na(c("a", NA, "b"))
```
```{r}
flights |> 
  filter(is.na(dep_time))
```
```{r}
flights |> 
  filter(month == 1, day == 1) |> 
  arrange(dep_time)
```
```{r}
flights |> 
  filter(month == 1, day == 1) |> 
  arrange(desc(is.na(dep_time)), dep_time)
```
### 12.2.4 Exercises
```{r}
flights |> 
  mutate(
    na_dep_time = is.na(dep_time),
    na_sched_dep_time = is.na(sched_dep_time),
    na_dep_delay = is.na(dep_delay)
  ) |> 
  count(na_dep_time & na_sched_dep_time & na_dep_delay )
```
## 12.3 Boolean algebra
### 12.3.1 Missing values
```{r}
df <- tibble(x = c(TRUE, FALSE, NA))
```
```{r}
df |> 
  mutate(
    and = x & NA,
    or = x | NA
  )
```
### 12.3.2 Order of operations
```{r}
flights |> 
  filter(month == 11 | month == 12)
```
```{r}
flights |> 
  filter(month == 11 | 12)
```
```{r}
flights |> 
  mutate(
    nov = month == 11,
    final = nov | 12,
    .keep = "used"
  )
```
### 12.3.3 %in%
```{r}
1:12 %in% c(1, 5, 11)
```
```{r}
letters[1:10] %in% c("a", "e", "i", "o", "u")
```
```{r}
flights |> 
  filter(month %in% c(11, 12))
```
```{r}
c(1, 2, NA) == NA
```
```{r}
c(1, 2, NA) %in% NA
```
```{r}
flights |> 
  filter(dep_time %in% c(NA, 0800))
```
### 12.3.4 Exercises
```{r}
flights |> 
  filter(is.na(arr_delay) & !is.na(dep_delay))
```
```{r}
flights |> 
  filter(
    !is.na(arr_time) & 
      !is.na(sched_arr_time) & 
      is.na(arr_delay))
```
```{r}
flights |> 
  filter(is.na(dep_time)) |>  
  summarise(across(everything(), ~ sum(is.na(.)))) |>  
  select(where(~ . > 0))
```
```{r}
canceled_flights <- flights |> 
  mutate(cancelled = is.na(dep_time)) |> 
  group_by(year, month, day) |> 
  summarise(
    canceled_flights = sum(cancelled),
    total_flights = n(),
    canceled_prop = mean(cancelled),
    .groups = "drop"
  )
```
```{r}
canceled_flights
```
## 12.4 Summaries
### 12.4.1 Logical summaries
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    all_delayed = all(dep_delay <= 60, na.rm = TRUE),
    any_long_delay = any(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```
### 12.4.2 Numeric summaries of logical vectors
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    proportion_delayed = mean(dep_delay < 60, na.rm =  TRUE),
    count_long_delay = sum(arr_delay >= 300, na.rm = TRUE),
    .groups = "drop"
  )
```
### 12.4.3 Logical subsetting
```{r}
flights |> 
  filter(arr_delay > 0) |> 
  group_by(year, month, day) |> 
  summarize(
    behind = mean(arr_delay),
    n = n(),
    .groups = "drop"
  )
```
```{r}
flights |> 
  group_by(year, month, day) |> 
  summarize(
    behind = mean(arr_delay[arr_delay > 0], na.rm = TRUE),
    ahead = mean(arr_delay[arr_delay < 0], na.rm = TRUE),
    n = n(),
    .groups = "drop"
  )
```
## 12.5 Conditional transformations
### 12.5.1 if_else()
```{r}
x <- c(-3:3, NA)
```
```{r}
if_else(x > 0, "+ve", "-ve")
```
```{r}
if_else(x < 0, -x, x)
```
```{r}
x1 <- c(NA, 1, 2, NA)
```
```{r}
y1 <- c(3, NA, 3, 6)
```
```{r}
if_else(is.na(x1), y1, x1)
```
```{r}
if_else(x == 0, "0", if_else(x < 0, "-ve", "+ve"), "???")
```
### 12.5.2 case_when()
```{r}
x <- c(-3:3, NA)
```
```{r}
case_when(
  x == 0 ~ "0",
  x < 0 ~ "-ve",
  x > 0 ~" +ve",
  is.na(x) ~ "???"
)
```
```{r}
case_when(
  x < 0 ~ "-ve",
  x > 0 ~ "+ve"
)
```
```{r}
case_when(
  x < 0 ~ "-ve",
  x > 0 ~ "+ve",
  .default = "???"
)
```
```{r}
case_when(
  x > 0 ~ "+ve",
  x > 2 ~ "big"
)
```
```{r}
flights |> 
  mutate(
    status = case_when(
      is.na(arr_delay) ~ "cancelled",
      arr_delay < -30 ~ "very early",
      arr_delay < -15 ~ "early",
      abs(arr_delay) <= 15 ~ "on time",
      arr_delay < 60 ~ "late",
      arr_delay < Inf ~ "very late"
    )
  )
```
### 12.5.3 Compatible types
```{r}
if_else(TRUE, "a", 1)
```
```{r}
case_when(
  x < -1 ~ TRUE,
  x > 0 ~ now()
)
```
### 12.5.4 Exercises
```{r}
x <- c(0:20)
```
```{r}
case_when(
  x %% 2 == 0 ~ "even",
  x %% 2 != 0 ~ "odd"
)
```
```{r}
x <- c("Monday", "Saturday", "Wednesday")
```

```{r}
if_else(
  x == "Saturday", "Weekend", "Weekday"
)
```

```{r}
x <- c(-5, 3, -8, 0, 7)
```

```{r}
abs_x <- if_else(x < 0, -x, x)
```
```{r}
abs_x
```




